A fast algorithm for global code motion of congruent instructions

### 2017-11-13 cgo/paper.tex
move sinking in legality

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a fast \gcm{} (\GCM{}) compiler optimization which schedules
congruent instructions across the program. We introduce data structures to help
exploit flow of global values in sparse representation.  Not only \GCM{}
improves performance it saves code size as well, it exposes more instruction
level parallelism in the basic-block to which instructions are moved, and it
enables other passes like loop invariant motion to remove redundancies. The cost
model to drive the code motion is based on liveness analysis on \SSA{}
representation such that the (virtual) register pressure does not increase
resulting in 2\% fewer spills on the \SPEC{} benchmark suite. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We also formalize why register pressure
reduces as a result of \gcm{}, and how \gcm{} increases instruction level
parallelism thereby enabling more \ooo{} execution on modern architectures.

We have implemented the pass in \LLVM{}. The experimental results show
performance gains of 8\% in 462.libquantum and 5.3\% in 471.omnetpp in \SPEC{}
benchmark suite. It also reduces compilation time by reducing number of
instructions to be processed by later compilation stages. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks.
\end{abstract}

@@

This paper describes a fast algorithm for \gcm{} (\GCM{}) of congruent
instructions \cite{briggs1997}, a technique that uses the information computed
for \PRE{} to detect identical computations but has a transformation phase whose
goal differs from \PRE{}: it removes identical computations from different
branches of execution.  These identical computations in different branches of
execution are not redundant computations at run-time and the number of run-time
computations is not reduced. It is not a redundancy elimination pass, and thus
it has different cost function and heuristics than \PRE{} or CSE. It is similar
to global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move computations. Code hoisting, for example, can reduce the critical
path length of execution in \ooo{} machines. As more instructions are available
at the hoisting point, the hardware has more independent instructions to
reorder. The following reduced example illustrates how hoisting can improve
performance by exposing more ILP.

@@

There has been a lot of work both in industry and academia related to code
hoisting/sinking, and in general global scheduling \cite{click1995global}. Some
relate code-hoisting to code-size optimization \cite{rosen1988global} and many
\cite{barany2013, shobaki2013} use global scheduling to improve
performance. Most of the recent work on global scheduling are done using integer
linear programming (ILP) which results in prohibitively high compile time which
is not suitable for industrial compilers. To the best of our knowledge we have
not found any reference which explored \gcm{} with a cost-model to reduce
register-pressure. A part of our implementation (aggressive code hoisting) is
already merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item introducing an analogue of $\phi$ nodes on the inverse CFG which allows
  faster tracking of anticipable instructions;
\item a cost model to reduce interference and hence, reduce spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

\section{Related Work}

There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic-blocks: the instructions of two sibling basic-blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalable.

@@

% TODO: References
% Sink instructions
% Ranking of instructions from Zadeck
% Liveness analysis from Ramakrishna
% Experimental results: Register spills, compile time, run time.

\section{Global code motion}

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item find candidates (congruent instructions) suitable for \gcm{}
  (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}),
\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
\end{itemize}

\section{Factoring reachability of expressions}
Our algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
being structural properties of a graph, do not change during code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it is the basic block on which the basic
block containing I is control dependent.

So we introduce a data structure ($\chi$ nodes) to keep track of values flowing
out of a basic block with more than one successor. This approach allows us to
hoist instructions to a basic block with more than two successors. The $\chi$
node in a basic block $B$ has CHIArgs with three entries: a) the instruction (I)
CHIArg refers to, b) the basic block $Dest$ which is post-dominated by the basic
block containing I, and c) the GVN of I.

@@

For that, we need to prove that on all execution paths from the initial position
of an instruction to its destination, the side effects appear in the same
order. In general, we cannot introduce a new computation along any path of
execution without checking for undefined behavior \cite{undef}. It is also
necessary to check if there are indirect branch targets, e.g., landing pad, case
statements, goto labels, etc., along the path because it becomes difficult to
prove safety checks in those cases.

Scalars are the easiest to hoist because we only need to check for availability
of operands. For sinking the scalars we need to make sure that there are no uses
of the instruction along the path, For instructions with memory references like
loads, stores and calls, Memory SSA allows efficient way of iterating over the
use-def chains of memory references on a factored graph. Hoisting loads/stores
across calls also requires precise analysis of all the memory addresses accessed
by the call. The current implementation being an intraprocedural pass, cannot
schedule aggressively across calls. In the presence of pure calls, loads can be
hoisted but stores can't. Also, if a call throws exceptions, or if it may not
return, nothing can be scheduled across that call.

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \GCM{} is profitable.  That
takes into account the impact \GCM{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance but
at the same time to reduce code size as much as possible. Sometimes, hoisting is
not upward-safe \cite{click1995global}, e.g., if the expressions are in a
landing pad, sinking of those expressions may reduce code size. Following cost
models are implemented:

@@

After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
expressions will reduce the register pressure because two registers will be
freed at the insertion point but only one register will be required to assign
the definition of hoisted instruction $a0$. We have reduced performance
regressions with this cost model even if LLVM has a live-range splitting
\cite{cooper1998live} optimization.

Sinking may also reduce the register pressure in some cases, e.g., when the use
operands are not kills. For sinking, higher ranked expressions would be sunk
first. And it would be illegal to sink higher ranked identical expressions if
they are not fully anticipable in the common post-dominator. For example:

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not fully anticipable in B3. So without knowing the ability to sink
of `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the other hand
(d1, d2) can safely be sunk because their operands are readily available at the
sink point, i.e., B3. It should also be noted that, just because the expressions
are identical and operands are available, it still requires a unique
post-dominating PHI to use the exact same values to be legally sinkable.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a fast \gcm{} (\GCM{}) compiler optimization which schedules
congruent instructions across the program. We introduce data structures to help
exploit flow of global values in sparse representation.  Not only \GCM{}
improves performance it saves code size as well, it exposes more instruction
level parallelism in the basic-block to which instructions are moved, and it
enables other compiler optimizations like loop invariant motion to remove
redundancies. The cost model to drive the code motion is based on liveness
analysis on \SSA{} representation such that the (virtual) register pressure does
not increase resulting in 2\% fewer spills on the \SPEC{} benchmark
suite. \GCM{} is an optimistic algorithm in the sense that it considers all
congruent instructions in a function as potential candidates. We also formalize
why register pressure reduces as a result of \gcm{}, and how \gcm{} increases
instruction level parallelism thereby enabling more \ooo{} execution on modern
architectures.

We have implemented the pass in \LLVM{}. The experimental results show
performance gains of 8\% in 462.libquantum and 5.3\% in 471.omnetpp from the
\SPEC{} benchmark suite. It also reduces compilation time by reducing number of
instructions to be processed by later compilation stages. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks.
\end{abstract}

@@

This paper describes a fast algorithm for \gcm{} (\GCM{}) of congruent
instructions \cite{briggs1997}, a technique that uses the information computed
for \PRE{} to detect identical computations but has a transformation phase whose
goal differs from \PRE{}: it removes identical computations from different
branches of execution.  These identical computations in different branches of
execution are not redundant computations at run-time and the number of run-time
computations is not reduced. It is not a redundancy elimination pass, and thus
it has different cost function and heuristics than \PRE{} or CSE. It is similar
to global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move dynamic computations. Code hoisting, for example, can reduce the
critical path length of execution in \ooo{} machines. As more instructions are
available at the hoisting point, the hardware has more independent instructions
to reorder. The following reduced example illustrates how hoisting can improve
performance by exposing more ILP.

@@

The \GCM{} pass can be broadly divided into the following:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item factor the reachability of expressions (Section~\ref{sec:chi}).
\item find candidates (congruent instructions) suitable for \gcm{}
  (Section~\ref{subsec:optimistic}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}),
\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
\end{itemize}

There has been a lot of work both in industry and academia related to code
hoisting/sinking, and in general global scheduling \cite{click1995global}. Some
relate code-hoisting to code-size optimization \cite{rosen1988global} and many
\cite{barany2013, shobaki2013} use global scheduling to improve
performance. Most of the recent work on global scheduling are done using integer
linear programming (ILP) which results in prohibitively high compile time which
is not suitable for industrial compilers. To the best of our knowledge we have
not found any reference which explored \gcm{} with a cost-model to reduce
register-pressure. A part of our implementation (aggressive code hoisting) is
already merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item introducing an analogue of $\phi$ nodes which allows faster tracking of
  anticipable instructions;
\item a cost model to reduce register spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

\section{Related Work}
There are a lot of bug reports in \GCC{} and \LLVM{} bugzillas
\cite{GCCCodeHoistingBugs, LLVMCodeHoistingBugs}, showing the interest in having
a more powerful code hoist transform.  The current \LLVM{} implementation of
code hoisting in SimplifyCFG.cpp is very limited to hoisting from identical
basic-blocks: the instructions of two sibling basic-blocks are read at the same
time, and all the instructions of the blocks are hoisted to the common parent
block as long as the compiler is able to prove that the instructions are
equivalent.  This implementation does not allow for an easy extension: first in
terms of compilation time overhead the implementation is quadratic in number of
instructions to bisimulate and second, the equivalence of instructions is
computed by comparing the operands which is neither general nor scalable.

@@

% TODO: References
% Sink instructions
% Ranking of instructions from Zadeck
% Liveness analysis from Ramakrishna
% Experimental results: Register spills, compile time, run time.
\section{Global Code Motion of congruent computations}
\label{sec:gcm}
The \GCM{} uses the following data structures ($\chi$ and $\Phi$ nodes), that
are introduced below. They allow us to factor the reachability of values in an
efficient way. Together with liveness analysis using the DJ-Graph \cite{das2012}
and Memory SSA, the entire implementation of \gcm{} only uses sparse
representations for efficient query and update.

\subsection{Factoring reachability of expressions}
\label{sec:chi}
Our algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers (DF) and the post-dominance frontiers (PDF) of the CFG. These points,
being structural properties of a graph, do not change during the code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it is the basic block on which the basic
block containing I is control dependent.

So we introduce a data structure ($\chi$ nodes) to keep track of values flowing
out of a basic block with more than one successor. This approach allows us to
hoist instructions to a basic block with multiple successors. The $\chi$ node in
a basic block $B$ has CHIArgs with three entries: a) the instruction (I) CHIArg
refers to, b) the basic block $Dest$ which is post-dominated by the basic block
containing I, and c) the GVN of I.

@@

We need to prove that on all execution paths from the initial position of an
instruction to its destination, the side effects appear in the same order. In
general, we cannot introduce a new computation along any path of execution
without checking for undefined behavior \cite{undef}. It is also necessary to
check if there are indirect branch targets, e.g., landing pad, case statements,
goto labels, etc., along the path because it becomes difficult to prove safety
checks in those cases.

Scalars are the easiest to hoist because we only need to check for availability
of operands. For sinking the scalars we need to make sure that there are no uses
of the instruction along the path. For instructions with memory references like
loads, stores and calls, Memory SSA allows efficient way of iterating over the
use-def chains of memory references on a factored graph. Hoisting loads/stores
across calls also requires precise analysis of all the memory addresses accessed
by the call. The current implementation being an intraprocedural pass, cannot
schedule aggressively across calls. In the presence of pure calls, loads can be
hoisted but stores can't. Also, if a call throws exceptions, or if it may not
return, nothing can be scheduled across that call.

For sinking, higher ranked expressions would be sunk first. And it would
be illegal to sink higher ranked identical expressions if they are not fully
anticipable in the common post-dominator. For example:

In this example (c1, c2) or (d1, d2) are potential sinkable candidates. Since
(c1, c2) depend on i1 and i2 respectively which are also in their original basic
blocks, c1 and c2 are not fully anticipable in B3. So without knowing the
ability to sink `i1' and `i2' it would be illegal to sink (c1, c2) to B3. On the
other hand (d1, d2) can safely be sunk because their operands are readily
available at the sink point, i.e., B3. It should also be noted that, just
because the expressions are identical and operands are available, it still
requires a unique post-dominating $\Phi$ to use the exact same values to be
legally sinkable.

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \GCM{} is profitable.  That
takes into account the impact \GCM{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance but
reduce register spills. We implement a cost model to reduce register pressure.

@@

After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ is placed in
$B0$. Since $b$ and $c$ are killed in both $a0$ and $a1$, hoisting the
expressions will reduce the register pressure because two registers will be
freed at the insertion point but only one register will be required to assign
the definition of hoisted instruction $a0$. We have reduced performance
regressions with this cost model even if LLVM has a live-range splitting
\cite{cooper1998live} optimization. Sinking may also reduce the register
pressure when the use operands are not kills.
